# Proteomics Data Viz

## Visualizing Proteomics data with ggplot2


```{r, include = FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=3, fig.align="center")
```

Last session we worked on Cross-linking Mass Spectrometry Data. The data consisted of interaction between 300 yeast nuclear proteins. 
We also learned about concepts behind `Grammer of Graphics` and plotting using `ggplot2`. 

In this tutorial, we will be working on the data from Yeast Nuclear Protein interaction study using Cross-linking Mass Spectrometry. 

```{r message=F, warning=F}
library(tidyverse)
```

We continue with the Cross-linking Proteomics dataset from Cytoscape tutorial.
The data was in excel, therefore we first converted it into `Comma separated file (csv)`  format.

### Optional: Reading excel file into R {.extra}
If you want to load the excel files files directly to `R` then you can use another library `readxl`. 

```{r message=F, warning=F}
library(readxl)
nuclear_xl_ms_excel <- readxl::read_excel("r-intro-files/Nuclear_XL_MS.xlsx")
head(nuclear_xl_ms_excel)
```


## Exploring the data
```{r message=F, warning=F}
library(readxl)
nuclear_xl_ms <- read_csv("r-intro-files/Nuclear_XL_MS.csv")
head(nuclear_xl_ms)
```

Now lets examine the dataset with two base R functions `str` and `summary`

```{r message=F, warning=F}
str(nuclear_xl_ms)
summary(nuclear_xl_ms)
```
While the data in `PPINovelty` and `PPIEvidenceInfoGroup` are characters/strings, they can also be thought of as categorical. 

We will now look into the unique values for the PPINovelty and PPIEvidenceInfoGroup columns

```{r message=F, warning=F}

nuclear_xl_ms$PPINovelty %>% unique()
```
```{r message=F, warning=F}
nuclear_xl_ms$PPIEvidenceInfoGroup %>% unique()
```


